---
title: "Homework 2"
author: "Sarah Haberlack"
date: "2023-09-30"
output: github_document
---

## Problem 1

# About the Data
"cpi" shows us the value of the consumer price index on the associated date.
"GDP" shows us the GDP on the associated date.
"pols-month" shows us the number of national politicians who are D or R on the associated date.
"recessions" shows us the start and end dates of a recession.
"snp" shows us the closing values of the S&P stock index on the associated date.
"unemployment" shows us the percentage of unemployment of an associated month.

# Cleaning pols_month 

Upload pols-month
```{r}
library(tidyverse)
pols_month = read.csv("data/fivethirtyeight_datasets/pols-month.csv") |>
  janitor::clean_names()
```

Break up variable "mon" into integer variables year, month, and day; replace month number with month name, create a president variable taking values of gop and dem
```{r}
pols_month = separate(pols_month, mon, c("year", "month", "day"), "-")|>
  mutate(month = case_match(
    month,
    "01" ~ "January",
    "02" ~ "February",
    "03" ~ "March",
    "04" ~ "April",
    "05" ~ "May",
    "06" ~ "June",
    "07" ~ "July",
    "08" ~ "August",
    "09" ~ "September",
    "10" ~ "October",
    "11" ~ "November",
    "12" ~ "December")) |>
  
  mutate(president = ifelse(prez_gop == 1, "Republican", "Democrat")) |>
        select(-prez_gop & -prez_dem & -day) 
```

## Problem 2

Cleaning mr trash wheel, professor trash wheel and Gwynnda:
```{r}
mrtrash_df = readxl::read_excel("data/202309 Trash Wheel Collection Data.xlsx", sheet = 1, range = "A2:N586") |>
janitor::clean_names() |>
mutate(homes_pwrd = weight_tons*(50/3), wheel = "mister") |>
  select(wheel, everything(), -year)

proftrash_df = readxl::read_excel("data/202309 Trash Wheel Collection Data.xlsx", sheet = 2, range = "A2:M108") |>
janitor::clean_names() |>
mutate(homes_pwrd = weight_tons*(50/3), wheel = "prof") |>
  select(wheel, everything(), -year)


gwynnda_df = readxl::read_excel("data/202309 Trash Wheel Collection Data.xlsx", sheet = 4, range = "A2:L157") |>
janitor::clean_names() |>
mutate(homes_pwrd = weight_tons*(50/3), wheel = "gwynnda") |>
  select(wheel, everything(), -year)
```

# Combine datasets
```{r}
comb_df = bind_rows(mrtrash_df, proftrash_df, gwynnda_df) |>
view()
```

There are 835 observations in the combined dataset. Key variables include `wheel`, to indicate whether the dumpster came from Mr. Trashwheel, Professor Trashwheel, or Gwynnda Trashwheel; `date` to indicate which day this dumpster was filled, variables that indicate `weight_tons`, `volume_cubic_yards`, assorted trash materials that were collected, and `homes_pwrd` which indicates average number of home-days powered per observation. 

The total weight of trash collected by Professor Trash Wheel is `r sum(proftrash_df$weight_tons)` tons.

The total number of cigarette butts collected by Gwynnda in July of 2021 is XXX.

## Problem 3

# Clean MCI data
```{r}
library(tidyverse)
baseline_df = read.csv("data/data_mci/MCI_baseline.csv", skip = 1) |>
  janitor::clean_names() |>
  mutate(
    sex = 
      case_match(
        sex, 
        1 ~ "male", 
        0 ~ "female"),
    sex = as.factor(sex)) |>
  mutate(
    apoe4 =
        case_match(
          apoe4,
          1 ~ "carrier",
          0 ~ "non-carrier"),
    apoe4 = as.factor(apoe4)) |>
  rename(education_years = education, baseline_age = current_age) |>
  mutate(
    age_at_onset = as.numeric(age_at_onset)
  ) |> 
  mutate(
  age_diff = age_at_onset - baseline_age) |>
  filter(age_diff > 0 | is.na(age_diff)) |>
view()
```
At the beginning of the import process, we had to remove the first row (`skip = 1` within the `read.csv` command). This row served as a longer explanation for the actual columns beneath them. Without removing it, the column names are absurdly long, and it looks as though the true column row is the first observation. Because that column context is removed, we must now translate the numeric code used for `sex` and `apoe4` into character variables of male/female and carrier/non-carrier. This makes the data legible to those who are looking at it through our dataset for the first time. Next, we must remove all study participants who never developed an MCI. These would be the observations for `age_at_onset` - `basline_age` is less than 0. I created a new variable `age_diff` and filtered out observations for which `age_diff` was a negative number. Finally, for clarity's sake, I extended the `education` column name to `education_years`, another column name that became less self-explanatory when the lengthier row 1 was deleted. 

`r count(baseline_df)` participants were recruited to the study. 

Of these, `r count(baseline_df, is.na(age_at_onset))` participants developed MCI. 

The average baseline age was `r mean(baseline_df$baseline_age)` years.

`r ((count(filter(baseline_df, sex == "female" & apoe4 == "carrier")))/(count(filter(baseline_df, sex == "female"))))*100`% of women in the study are APOE4 carriers.