---
title: "Homework 2"
author: "Sarah Haberlack"
date: "2023-09-30"
output: github_document
---

## Problem 1

# About the Data
"cpi" shows us the value of the consumer price index on the associated date.
"GDP" shows us the GDP on the associated date.
"pols-month" shows us the number of national politicians who are D or R on the associated date.
"recessions" shows us the start and end dates of a recession.
"snp" shows us the closing values of the S&P stock index on the associated date.
"unemployment" shows us the percentage of unemployment of an associated month.

# Cleaning pols_month 

Upload pols-month
```{r}
library(tidyverse)
pols_month = read.csv("data/fivethirtyeight_datasets/pols-month.csv") |>
  janitor::clean_names()
```

Break up variable "mon" into integer variables year, month, and day; replace month number with month name, create a president variable taking values of gop and dem
```{r}
pols_month = separate(pols_month, mon, c("year", "month", "day"), "-")|>
  mutate(month = case_match(
    month,
    "01" ~ "January",
    "02" ~ "February",
    "03" ~ "March",
    "04" ~ "April",
    "05" ~ "May",
    "06" ~ "June",
    "07" ~ "July",
    "08" ~ "August",
    "09" ~ "September",
    "10" ~ "October",
    "11" ~ "November",
    "12" ~ "December"), 
    year = as.numeric(year), month = as.numeric(month)) |>
  mutate(president = ifelse(prez_gop == 1, "Republican", "Democrat")) |>
        select(-prez_gop & -prez_dem & -day) 
```

We also clean the 538 `snp` data, which contains information related to Standard & Poor’s stock market index.

```{r clean_538_snp}
snp = 
  read_csv("./data/fivethirtyeight_datasets/snp.csv",
  col_types = cols(date = col_date(format = "%m/%d/%y"))) |>
  separate(date, into = c("year", "month", "day"), convert = TRUE) |>
  arrange(year, month) |>
  mutate(
    year = if_else(year > 2023, year - 100, year)) |>
  left_join(x = _, y = pols_month) |>
  select(year, month, close) 
```

Finally, we tidy the `unemployment` data so that it can be merged with the `pols` and `snp` datasets.

```{r clean_538_unemp}
unemployment = 
  read_csv("./data/fivethirtyeight_datasets/unemployment.csv") |>
  rename(year = Year) |>
  pivot_longer(
    Jan:Dec, 
    names_to = "month_abb",
    values_to = "unemployment"
  ) |> 
  left_join(x = _, y = pols_month) |> 
  select(year, month, unemployment)
```



## Problem 2

Cleaning mr trash wheel, professor trash wheel and Gwynnda:
```{r}
mrtrash_df = readxl::read_excel("data/202309 Trash Wheel Collection Data.xlsx", sheet = 1, range = "A2:N586") |>
janitor::clean_names() |>
mutate(homes_pwrd = weight_tons*(50/3), wheel = "mister", 
    year = as.numeric(year)) |>
  select(wheel, everything())

proftrash_df = readxl::read_excel("data/202309 Trash Wheel Collection Data.xlsx", sheet = 2, range = "A2:M108") |>
janitor::clean_names() |>
mutate(homes_pwrd = weight_tons*(50/3), wheel = "prof") |>
  select(wheel, everything())


gwynnda_df = readxl::read_excel("data/202309 Trash Wheel Collection Data.xlsx", sheet = 4, range = "A2:L157") |>
janitor::clean_names() |>
mutate(homes_pwrd = weight_tons*(50/3), wheel = "gwynnda") |>
  select(wheel, everything())
```

# Combine datasets
```{r}
comb_df = bind_rows(mrtrash_df, proftrash_df, gwynnda_df) |>
view()
```

There are 835 observations in the combined dataset. Key variables include `wheel`, to indicate whether the dumpster came from Mr. Trashwheel, Professor Trashwheel, or Gwynnda Trashwheel; `date` to indicate which day this dumpster was filled, variables that indicate `weight_tons`, `volume_cubic_yards`, assorted trash materials that were collected, and `homes_pwrd` which indicates average number of home-days powered per observation. 

The total weight of trash collected by Professor Trash Wheel is `r sum(proftrash_df$weight_tons)` tons.

The total number of cigarette butts collected by Gwynnda in July of 2021 is `r gwcigs = gwynnda_df |> filter(year == 2021 & month == "July")
  sum(gwcigs$cigarette_butts)`.

## Problem 3

# Clean MCI data
```{r}
baseline_df = read.csv("data/data_mci/MCI_baseline.csv", skip = 1) |>
  janitor::clean_names() |>
  mutate(
    sex = 
      case_match(
        sex, 
        1 ~ "male", 
        0 ~ "female"),
    sex = as.factor(sex)) |>
  mutate(
    apoe4 =
        case_match(
          apoe4,
          1 ~ "carrier",
          0 ~ "non-carrier"),
    apoe4 = as.factor(apoe4)) |>
  rename(education_years = education, baseline_age = current_age) |>
  mutate(
    age_at_onset = as.numeric(age_at_onset)
  ) |> 
  mutate(
  age_diff = age_at_onset - baseline_age) |>
  filter(age_diff > 0 | is.na(age_diff)) |>
view()
```
At the beginning of the import process, we had to remove the first row (`skip = 1` within the `read.csv` command). This row served as a longer explanation for the actual columns beneath them. Without removing it, the column names are absurdly long, and it looks as though the true column row is the first observation. Because that column context is removed, we must now translate the numeric code used for `sex` and `apoe4` into character variables of male/female and carrier/non-carrier. This makes the data legible to those who are looking at it through our dataset for the first time. Next, we must remove all study participants who never developed an MCI. These would be the observations for `age_at_onset - baseline_age` is less than 0. I created a new variable `age_diff` and filtered out observations for which `age_diff` was a negative number. Finally, for clarity's sake, I extended the `education` column name to `education_years`, another column name that became less self-explanatory when the lengthier row 1 was deleted. 

`r count(baseline_df)` participants were recruited to the study. 

Of these, 93 participants developed MCI.

The average baseline age was `r mean(baseline_df$baseline_age)` years.

`r ((count(filter(baseline_df, sex == "female" & apoe4 == "carrier")))/(count(filter(baseline_df, sex == "female"))))*100`% of women in the study are APOE4 carriers.

# Import, clean, tidy biomarker dataset
To clean this dataset, I will need to use the information in the explanatory row (which I will end up skipping) to inform the more useful variable names I create. These data represent the amyloid beta 42/40 ratio levels at timestamps after the baseline visits. I will change my variable names to represent this. I changed `study_id` to `id` so that this dataset can merge with `baseline_df`. I changed the amyloid ratio levels to numeric column types in case we need to do math with them later.
```{r}
amyloid_df = read.csv("data/data_mci/MCI_amyloid.csv", skip = 1) |>
  janitor::clean_names() |>
  rename(id = study_id, amyloid_baseline = baseline, amyloid_year_2 = time_2, amyloid_year_4 = time_4, amyloid_year_6 = time_6, amyloid_year_8 = time_8) |>
  mutate(
    amyloid_baseline = as.numeric(amyloid_baseline), amyloid_year_2 = as.numeric(amyloid_year_2), amyloid_year_4 = as.numeric(amyloid_year_4), amyloid_year_6 = as.numeric(amyloid_year_6), amyloid_year_8 = as.numeric(amyloid_year_8)
  )
```


```{r}
anti_join(baseline_df, amyloid_df, by = "id")
```
There are 8 participants who appear in only the baseline dataset, but not the amyloid dataset. There are no participants who appear in the amyloid dataset but not the baseline dataset. This could be because it is much easier to get data about baseline characteristics of a participant as opposed to a biomarker, which requires taking a blood sample--a much more precarious situation which may not always be successful.

# Inner join baseline and amyloid datasets

```{r}
fulldf = inner_join(baseline_df, amyloid_df) |>
  view()
```

The resulting dataset has `r count(fulldf)` participants. The mean baseline age is `r mean(fulldf$baseline_age, na.rm = TRUE)`.

```{r}
write.csv(fulldf, "data/alzheimers_trajectory.csv")
```

