---
title: "Homework 2"
author: "Sarah Haberlack"
date: "2023-09-30"
output: github_document
---

## Problem 1

# About the Data
"cpi" shows us the value of the consumer price index on the associated date.
"GDP" shows us the GDP on the associated date.
"pols-month" shows us the number of national politicians who are D or R on the associated date.
"recessions" shows us the start and end dates of a recession.
"snp" shows us the closing values of the S&P stock index on the associated date.
"unemployment" shows us the percentage of unemployment of an associated month.

# Cleaning pols_month 

Upload pols-month
```{r}
library(tidyverse)
pols_month = read.csv("data/fivethirtyeight_datasets/pols-month.csv") |>
  janitor::clean_names()
```

Break up variable "mon" into integer variables year, month, and day; replace month number with month name, create a president variable taking values of gop and dem
```{r}
pols_month = separate(pols_month, mon, c("year", "month", "day"), "-")|>
  mutate(month = case_match(
    month,
    "01" ~ "January",
    "02" ~ "February",
    "03" ~ "March",
    "04" ~ "April",
    "05" ~ "May",
    "06" ~ "June",
    "07" ~ "July",
    "08" ~ "August",
    "09" ~ "September",
    "10" ~ "October",
    "11" ~ "November",
    "12" ~ "December")) |>
  
  mutate(president = ifelse(prez_gop == 1, "Republican", "Democrat")) |>
        select(-prez_gop & -prez_dem & -day) 
```